Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-29 demystify.tex
updates

\paragraph{What else?} Make decision (\autoref{sec:accept-postpone-decline})! accept, reject, or defer the offers (if you have multiple options). Ask to meet with potential advisers (e.g., through Open House or separately) and even their students. Ask about computer equipment and software, office space, and other resources; in many case these will be provided for free by your adviser or department (\autoref{sec:buying-equipment}).

Also, do not forget to update and thank your LoR writters and others who have supported you through this process.

\paragraph{If you do not get admitted} If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice (\autoref{sec:improve-your-chance}).

You can consider applying to MS programs, which are typically easier to get in; but you likely need to pay.  If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.

@@

\begin{commentbox}[Vu:]
  TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At GMU, and at most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically default for CS PhD programs. When you apply to be a full-time student,  state that you need financial assistant. It is common that the PhD committee will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

@@

Even if you have other funding and do not need TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for research career where you often give talks and tell people about your work. GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as GTA or even sometimes as a lecturer.  This is a good opportunity for students to get teaching experience and also get paid more (as a lecturer).

@@

\begin{commentbox}[Vu:]
  If you got recruited an offered an RA by a prof., you will likely get admitted.  For example, if a prof., even if not in PhD adcom, wants to fund you, the PhD adcom will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, government).
Fellowships are typically competitive and generous, giving pretty much all benefits tuition/insurance that a TA/RA has.  They might even give higher stipend (including summer) and open doors for job opportunities such as internships.

\textbf{How to get Fellowship?}   Many schools provide fellowships to attract students. You likely will not need to do anything and the adcom will recommend such fellowships to strong students. Some schools automatically offer a fellowship to all accepted students, while others only award it to a limited number of admitted students, such as the top percentile.

@@

\begin{commentbox}[Vu:]
  PhD applicants at GMU are automatically eligible for a \emph{Presidential Fellowship}.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you do not depend on any prof. or TA).  PhD adcom members nominate applicants for this fellowship and the whole committee will decide.
\end{commentbox}

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizen and permanent residents, for underrepresented minority (URM) students (\autoref{sec:urm}), and for students from specific countries. There are many fellowships available, so you should look for them and apply.

@@

\section{Choosing a University}

@@

You can learn about CS programs and research interests of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

\section{Choosing an Adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with.  But how do you find such a person?

@@

Here is some general advice to find a professor.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research that professor and contact them (\autoref{sec:contact}).

@@

You can also ask if you can do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I do this with multiple students, many of whom are undergrads. Many will drop because they find they don't like my research, but some will stay and become my PhD students.

@@

  In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
  \tcblower
  \textbf{Vu}: \autoref{sec:contact} shows how to contact a professor for research opportunities. Many probably will say no (or do not reply) as they do not have the bandwidth to take on random students, but some may say yes if they see potential fit.
\end{commentbox}

@@

\paragraph{Tenure-line and teaching faculty} Tenure-line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring PhD students.  They often have low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  At most universities, \textbf{only} tenure and tenured track faculty can serve as a formal adviser of PhD students. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

In contrast, \emph{teaching} (or instructional faculty or professor of practice) mainly focus on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. They can mentor PhD students but they typically cannot serve as a formal adviser of PhD students, i.e., they \emph{cannot graduate} PhD students.

@@

\paragraph{Titles} Faculty have titles such as assistant, associate, full, regardless if they are tenure line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching faculty or lecturer, also starts with assistant, then promoted to associated, and full.

Adjunct faculty is not full-time, e.g., they might be working in industry and teach a class or two for fun.  Emeritus means they are retired but still have some affiliation with the university.  Research faculty (or research scientist) are typically non-tenure line faculty who focus on research.  Due to their roles, adjunct, emeritus, and research faculty typically do not advise PhD students.

@@

The short answer is that tenure-track faculty such as assistant professors are likely young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard. They might also expect you to figure things out yourself, i.e., you need to be independent.  Some tenured faculty are also no longer active in research and more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students).

@@

Faculty often receive ``cold'' e-mails from prospective students. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So how to write an email that get our attention?

@@

\paragraph{Overqualified or Lack Interest}  This might be paradoxical and is the opposite of aiming too high. However if the adcom believes that you are likely to get admitted and go to a better program (i.e., they are just your ``safety'' schools), they might not be so enthusiastic to admit you and want to save the spot for someone else. Note though it does happen, for most good schools this is rare and you should not assume this is the case.

A related reason is that you did not show enough interests in the program.
For example, your SOP did not mention any faculty or research area at the program, you did not respond to professors interested in interviewing you (which might be considered unprofessional and burn bridges), or during the interview you show little interest or enthusiasm (because they are your safety schools) in the program or profs.

\paragraph{Red flags in Your Background} Plagiarism, cheating, or other academic dishonesty are \red{red} flags. Another one is if you have a history of jumping from one program to another (e.g., you have been in multiple different programs in short time). How would the reviewers know?  They might have contacts in other programs and heard about you, your LoRs might mention it.

>>>>

\paragraph{What else?} Make decision, accept, reject, or defer the offers  (\autoref{sec:accept-postpone-decline}). Ask to meet with potential advisers (e.g., through Open House or separately) and even their students. Ask about computer equipment and software, office space, and other resources; in many case these will be provided for free by your adviser or department (\autoref{sec:buying-equipment}).

Also, do not forget to update and thank your LoR writers and others who have supported you through this process.

\section{If you do not get admitted}\label{sec:not-accepted}

If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice (\autoref{sec:improve-your-chance}).

You can consider applying to MS programs, which are typically easier to get in; but you likely need to pay.  Sometimes, the university that rejected you for PhD might accept you for MS.
If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.

@@

\begin{commentbox}[Vu:]
  TA and RA at GMU have similar benefits in tuition waiving and insurance.  The college and department will set the rate for 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: really close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At GMU, and most good CS PhD programs, your GTA or GRA \emph{will always} come with full insurance. In fact, at GMU your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically the default for CS PhD programs. When you apply to be a full-time student,  state that you need financial assistant. Typically, adcom will either admit you and give you GTA, or reject you; i.e., we do not admit a student without supporting them.

@@

Even if you have other funding and do not need a TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for a research career where you often give talks and tell people about your work. GMU sometimes has classes that a more senior student can teach.  In that case, you will be paid as GTA or even sometimes as a lecturer.  This is a good opportunity for students to get teaching experience and also get paid more (as a lecturer).

@@

\begin{commentbox}[Vu:]
  If you got recruited an offered an RA by a prof., you will likely get admitted.  For example, if a prof., even if not in PhD adcom, wants to fund you, adcom will respect that decision and admit you (unless your application has many red flags).
\end{commentbox}

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, or government).
Fellowships are typically competitive and generous, giving pretty much all benefits tuition/insurance that a TA/RA has.  They might even give higher stipends (including summer) and open doors for job opportunities such as internships.

\textbf{How to get Fellowship?}   Many schools provide fellowships to attract students. You likely will not need to do anything and adcom will recommend such fellowships to strong students. Some schools automatically offer a fellowship to all accepted students, while others only award it to a limited number of admitted students, such as the top percentile.

@@

\begin{commentbox}[Vu:]
  PhD applicants at GMU are automatically eligible for a \emph{Presidential Fellowship}.  It is at least as good as GTA but the most important thing is that as a fellowship it is truly free money (i.e., you do not depend on any prof. or TA).  Adcom members nominate applicants for this fellowship and the whole committee will decide.
\end{commentbox}

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizens and permanent residents, for underrepresented minority (URM) students (\autoref{sec:urm}), and students from specific countries. There are many fellowships available, so you should look for them and apply.

@@

\section{Choosing a University}\label{sec:choosing-university}

@@

You can learn about CS programs and the research interests of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}. You will be very surprised to learn that a school that you didn't know much about can have very strong research in your topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty, e.g., who works on what, and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research findings (see \href{https://homes.cs.washington.edu/~mernst/advice/conferences-vs-journals.html}{why here}).}. %\autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

\section{Choosing an Adviser}
There is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with.  But how do you find such a person?

@@

Here is some general advice for finding a professor.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research that professor and contact them (\autoref{sec:contact}).

@@

You can also ask if you can do an independent study or research with a prof. This can be informal (no credit) and takes place during the summer or winter break.  For example, I do this with multiple students, many of whom are undergrads. Many will drop out because they find they don't like my research, but some will stay and become my PhD students.

@@

  In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
  \tcblower
  \textbf{Vu}: \autoref{sec:contact} shows how to contact a professor for research opportunities. Many probably will say no (or do not reply) as they do not have the bandwidth to take on random students, but some may say yes if they see a potential fit.
\end{commentbox}

@@

\paragraph{Tenure-line and teaching faculty} Tenure-line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, and mentoring PhD students.  They often have a low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  At most universities, \textbf{only} tenure and tenured track faculty can serve as a formal adviser of PhD students. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

In contrast, \emph{teaching} (or instructional faculty or professor of practice) mainly focuses on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities, e.g., they do not have to worry much about publishing papers or obtaining grants. They can mentor PhD students but they typically cannot serve as a formal adviser of PhD students, i.e., they \emph{cannot graduate} PhD students.

@@

\paragraph{Titles} Faculty have titles such as assistant, associate, full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching faculty or lecturer, also starts with assistant, then promoted to associated, and full prof.

Adjunct faculty is not full-time, e.g., they might be working in industry and teach a class or two for fun.  Emeritus means they are retired but still have some affiliation with the university.  Research faculty (or research scientists) are typically non-tenure line faculty who focus on research.  Due to their roles, adjunct, emeritus, and research faculty typically do not advise PhD students.

@@

The short answer is that tenure-track faculty such as assistant professors are likely young and active in research (they have to, to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard. They might also expect you to figure things out yourself, i.e., you need to be independent.  Some tenured faculty are also no longer active in research and are more involved with administrative responsibilities or with their startup companies (this means they will likely not take new students).

@@

Faculty often receive ``cold'' e-mails from prospective students. Most of the time, we ignore these emails, but in some rare occasions we do answer them. So how to write an email that gets our attention?

@@

\paragraph{Overqualified or Lack Interest}  This might be paradoxical and is the opposite of aiming too high. However if adcom believes that you are likely to get admitted and go to a better program (i.e., they are just your ``safety'' schools), they might not be so enthusiastic to admit you and want to save the spot for someone else. Note though it does happen, for most good schools this is rare and you should not assume this is the case.

A related reason is that you did not show enough interest in the program.
For example, your SOP did not mention any faculty or research area at the program, you did not respond to professors interested in interviewing you (which might be considered unprofessional and burn bridges), or during the interview you show little knowledge, interest, or enthusiasm (because they are your safety schools) in the program or profs.

\paragraph{Red flags in Your Background} Plagiarism, cheating, or other academic dishonesty are \red{red} flags. Another one is if you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). How would the reviewers know?  They might have contacts in other programs and heard about you, your LoRs might mention it.
